---
title: "Ball, strike, foul, and barrel values"
author: "Jason Willwerscheid"
date: "2/20/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Count-specific wOBA

I calculate expected wOBA conditional on the pitch count (the idea is described in Tango, Lichtman, and Dolphin). To account for the juicier balls of 2017 and 2019, I want these on a per-year basis, but for less frequent pitch counts like 3-0, the standard errors associated with the empirical means are a bit too large for my taste. Thus I make the additional assumption that year-to-year differences are constant across pitch counts. (I could have also assumed that the ratios are constant, but differences are easier to interpret.) As an example, I give the empirical and fitted values for 3-0 below: one can verify that the fitted wOBA values are all very reasonable given the empirical wOBAs and SEs:

```{r count_woba}
suppressMessages(library(tidyverse))

all_pitches <- readRDS("output/mlb_wprobs.rds")

# TODO: Technically, each PA should only count once; so if a PA stays at 3-2 for several
#   pitches, it shouldn't be counted more than once.

count_woba <- all_pitches %>%
  group_by(game_year, balls, strikes) %>%
  summarize(empirical_wOBA = mean(pa_woba_value),
            empirical_SE = sd(pa_woba_value) / n()) %>%
  ungroup() %>%
  arrange(balls, strikes, game_year)

lm_fit <- lm(pa_woba_value ~ factor(game_year) + factor(balls)*factor(strikes), 
             data = all_pitches)
lm_preds <- predict(lm_fit, newdata = count_woba)
count_woba <- count_woba %>%
  mutate(fitted_wOBA = lm_preds, 
         diff = fitted_wOBA - empirical_wOBA)

knitr::kable(count_woba %>% filter(balls == 3, strikes == 0), digits = 3)
```

The full table for 2019 is:
```{r count_woba2}
count_woba <- count_woba %>%
  rename(wOBA = fitted_wOBA) %>%
  select(game_year, balls, strikes, wOBA)
saveRDS(count_woba, "output/count_woba.rds")

knitr::kable(count_woba %>% filter(game_year == 2019), digits = 3)  
```

## wOBACON

I calculate expected wOBACON separately for balls that are barreled and those that aren't:

```{r wobacon}
wobacon <- all_pitches %>%
  filter(contact == 1, foul == 0) %>%
  group_by(game_year, barrel) %>%
  summarize(wOBACON = mean(pa_woba_value)) %>%
  ungroup() %>%
  spread(key = barrel, value = wOBACON) %>%
  rename(nonbrl_wOBACON = `0`, barrel_wOBACON = `1`)

knitr::kable(wobacon, digits = 3)
```

## wOBA values for pitch events

Finally, I calculate the value of a ball, strike, foul ball, non-barreled ball, and barreled ball conditional on the pitch count. 

```{r event_vals}
helper_df <- tibble(balls = 4, strikes = 0:2, wOBA = 1, dummy = 1) %>%
  bind_rows(tibble(strikes = 3, balls = 0:3, wOBA = 0, dummy = 1)) %>%
  full_join(tibble(game_year = 2015:2019, dummy = 1), by = "dummy") %>%
  select(-dummy)

event_vals <- count_woba %>%
  bind_rows(helper_df) 
event_vals <- event_vals %>%
  left_join(event_vals, by = "game_year", suffix = c("", "_next")) %>%
  filter((balls_next == balls + 1 & strikes_next == strikes)
         | (strikes_next == strikes + 1 & balls_next == balls)) %>%
  mutate(event = ifelse(strikes_next == strikes + 1, "strike", "ball"), 
         wOBA_change = wOBA_next - wOBA) %>%
  filter(balls < 4, strikes < 3) %>%
  select(game_year, balls, strikes, wOBA, event, wOBA_change) %>%
  spread(key = "event", value = "wOBA_change") %>%
  rename(strike_value = strike, ball_value = ball)
  
# Foul balls:
event_vals <- event_vals %>%
  mutate(foul_value = ifelse(strikes == 2, 0, strike_value))

# Balls in play:
event_vals <- event_vals %>%
  left_join(wobacon, by = "game_year") %>%
  mutate(nonbrl_value = nonbrl_wOBACON - wOBA,
         barrel_value = barrel_wOBACON - wOBA) %>%
  select(-wOBA, -nonbrl_wOBACON, -barrel_wOBACON)

all_pitches <- all_pitches %>%
  left_join(event_vals, by = c("game_year", "balls", "strikes"))
```

For example, let the count be 3-1 in 2019. This count is associated with an expected wOBA of 0.489. A strike will bring the count to 3-2, which has a wOBA of 0.392, so the value of a strike at 3-1 is -0.097. (And, due to the convenient assumption I made above, this value is constant from year to year.) A ball will result in a walk, which brings the wOBA up to 1: thus the value of a ball is 0.511. Since there are less than two strikes, a foul ball is equivalent to a strike. Finally, the value of hitting the ball into play depends on whether it's barreled or not: if so, the expected wOBACON is 1.279, so the value of the barrel is 0.790; if not, the expected wOBACON is 0.318, which results in a change of -0.172. At 3-1, then, putting the ball into play without barreling it is a much worse outcome than taking a strike!

```{r save_res}
saveRDS(all_pitches, "output/mlb_eventvals.rds")
```

