---
title: "Swing, called strike, contact, foul, and barrel probabilities"
author: "Jason Willwerscheid"
date: "2/20/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

My basic strategy is to condition all of my probabilities on pitch location and, when it makes sense, pitch count and pitch type.

I need five sets of probabilities:

* The probability that an average batter will swing at the pitch.
* The probability that the pitch will be called a strike, given that the batter doesn't swing.
* The probability that an average batter will make contact with the pitch, given that the batter does swing.
* The probability that the ball will be hit foul, given that the batter makes contact.
* The probability that the ball will be barreled, given that the batter makes contact.

I condition everything on pitch type except for called strike probabilities. I condition swing probabilities on pitch count, and I condition strike probabilities on whether or not there are already two strikes (on average, umpires are slightly more reluctant to call strike three than strikes one and two). All probabilities are estimated by binning and calculating empirically observed proportions. In a few cases, I needed to coarsify some of the bins to get reasonably smooth probability surfaces. In the plots below, the dashed lines indicate the [theoretical strike zone](https://www.baseballprospectus.com/news/article/37347/robo-strike-zone-not-simple-think/), and the line segment at `plate_z = 0` is home plate.


## Called strike probabilities

I first calculate
$$ p(\text{strike = 1} \mid \text{swing = 0, pitch location, I(strikes = 2)}) $$

```{r looks}
suppressMessages(library(tidyverse))

all_pitches <- readRDS("output/mlb_preproc.rds")

bins <- all_pitches %>%
  group_by(plate_x, plate_z, bin_ID, bin_type) %>%
  summarize() %>%
  ungroup()

plate_left <- -8.5 / 12
plate_right <- 8.5 / 12
sz_left <- plate_left - 2.94 / 2 / 12
sz_right <- plate_right + 2.94 / 2 / 12
sz_bot <- 15.53 / 12
sz_top <- 42.47 / 12

looks <- all_pitches %>%
  filter(swing == 0) %>%
  group_by(bin_ID, two_strikes = (strikes == 2)) %>%
  summarize(p_strike = sum(called_strike) / sum(swing == 0)) %>%
  ungroup()

ggplot(bins %>% left_join(looks, by = "bin_ID"), 
       aes(x = plate_x, y = plate_z, fill = p_strike)) + 
  geom_tile() +
  scale_fill_gradient2(high = "dark green") +
  theme_minimal() +
  geom_rect(aes(xmin = sz_left, xmax = sz_right, ymin = sz_bot, ymax = sz_top),
                col = "black", linetype = "dashed", alpha = 0) +
  geom_segment(aes(x = plate_left, xend = plate_right, y = 0, yend = 0)) +
  facet_wrap(~two_strikes, 
             labeller = labeller(two_strikes = c("FALSE" = "0-1 strikes", 
                                                 "TRUE" = "2 strikes")))
```


## Contact probabilities

Next I calculate
$$ p(\text{contact = 1} \mid \text{swing = 1, pitch location, pitch type}) $$
Recall that I flipped the $x$-axis for left-handed hitters, so there's no need to condition on handedness.

```{r contact}
contact <- all_pitches %>%
  group_by(bin_ID, pitch_type) %>%
  summarize(p_contact = ifelse(sum(swing) == 0, 0, sum(contact) / sum(swing))) %>%
  ungroup() 

ggplot(bins %>% left_join(contact, by = "bin_ID"), 
       aes(x = plate_x, y = plate_z, fill = p_contact)) + 
  geom_tile() +
  scale_fill_gradient2(high = "dark green") +
  theme_minimal() +
  geom_rect(aes(xmin = sz_left, xmax = sz_right, ymin = sz_bot, ymax = sz_top),
                col = "black", linetype = "dashed", alpha = 0) +
  geom_segment(aes(x = plate_left, xend = plate_right, y = 0, yend = 0)) +
  facet_wrap(~pitch_type, nrow = 3, ncol = 2)
```


## Foul probabilities

Next up is
$$ p(\text{foul = 1} \mid \text{contact = 1, pitch location, pitch type}) $$

```{r fouls}
fouls <- all_pitches %>%
  group_by(bin_ID, pitch_type) %>%
  summarize(p_foul = ifelse(sum(contact) == 0, 0, sum(foul) / sum(contact))) %>%
  ungroup() 

ggplot(bins %>% left_join(fouls, by = "bin_ID"), 
       aes(x = plate_x, y = plate_z, fill = p_foul)) + 
  geom_tile() +
  scale_fill_gradient2(high = "dark green") +
  theme_minimal() +
  geom_rect(aes(xmin = sz_left, xmax = sz_right, ymin = sz_bot, ymax = sz_top),
                col = "black", linetype = "dashed", alpha = 0) +
  geom_segment(aes(x = plate_left, xend = plate_right, y = 0, yend = 0)) +
  facet_wrap(~pitch_type, nrow = 3, ncol = 2)
```


## Barrel probabilities

Next is
$$ p(\text{barrel = 1} \mid \text{contact = 1, pitch location, pitch type}) $$

Since barrels are in general low-probability events, I needed to coarsify the bins for all pitch types.

```{r barrels}
barrels <- all_pitches %>%
  group_by(bin_ID, pitch_type) %>%
  summarize(n = sum(contact), 
            p_barrel = ifelse(sum(contact) == 0, 0, sum(barrel) / sum(contact))) %>%
  ungroup() 

sub_bins <- bins %>%
  mutate(coarse_x = ifelse((plate_x * 5) %% 2 == 0, plate_x - 0.2, plate_x),
         coarse_z = ifelse((plate_z * 5) %% 2 == 0, plate_z + 0.2, plate_z),
         coarse_bin = ifelse(bin_type == "octant",
                             sapply(str_split(bin_ID, ":"), `[[`, 1),
                             paste(coarse_x, coarse_z, sep = ","))) %>%
  select(bin_ID, coarse_bin) %>%
  distinct()

barrels <- barrels %>%
  left_join(sub_bins, by = c("bin_ID")) %>%
  group_by(coarse_bin, pitch_type) %>%
  mutate(p_barrel = sum(p_barrel * n) / sum(n)) %>%
  ungroup()

barrels <- barrels %>%
  select(-n, -coarse_bin)

ggplot(bins %>% left_join(barrels, by = "bin_ID"), 
       aes(x = plate_x, y = plate_z, fill = p_barrel)) + 
  geom_tile() +
  scale_fill_gradient2(high = "dark green") +
  theme_minimal() +
  geom_rect(aes(xmin = sz_left, xmax = sz_right, ymin = sz_bot, ymax = sz_top),
                col = "black", linetype = "dashed", alpha = 0) +
  geom_segment(aes(x = plate_left, xend = plate_right, y = 0, yend = 0)) +
  facet_wrap(~pitch_type, nrow = 3, ncol = 2)
```


## Swing probabilities

It remains to calculate

$$ p(\text{swing = 1} \mid \text{pitch count, pitch location, pitch type}) $$

```{r swings}
swings <- all_pitches %>%
  group_by(bin_ID, bin_type, pitch_type, balls, strikes) %>%
  summarize(n = n(), p_swing = sum(swing) / n()) %>%
  ungroup

# Coarsify bins for 3-0, 3-1, and 2-0:
swings <- swings %>%
  left_join(sub_bins, by = c("bin_ID")) %>%
  group_by(coarse_bin, pitch_type, balls, strikes) %>%
  mutate(p_swing = ifelse(balls - strikes > 1, 
                          sum(p_swing * n) / sum(n), p_swing)) %>%
  ungroup()

# Coarsify again for off-speed pitches throw at 3-0:
swings <- swings %>%
  group_by(bin_type, pitch_type, balls, strikes) %>%
  mutate(p_swing = ifelse(bin_type == "coord" & balls == 3 & strikes == 0 &
                            pitch_type %in% c("CU", "CH", "SL"),
                          sum(p_swing * n) / sum(n), p_swing)) %>%
  ungroup() 

swings <- swings %>%
  select(-n, -bin_type, -coarse_bin)
  
for (next_type in c("FF", "FT", "SL", "CH", "CU")) {
  plot_df <- bins %>% 
    left_join(swings, by = "bin_ID") %>% 
    filter(pitch_type == next_type) %>%
    mutate(pitch_count = paste(balls, strikes, sep = "-"))
  
  plt <- ggplot(plot_df, aes(x = plate_x, y = plate_z, fill = p_swing)) +
    geom_tile() +
    scale_fill_gradient2(high = "dark green") +
    facet_wrap(~pitch_count, nrow = 4, ncol = 3) +
    geom_rect(aes(xmin = sz_left, xmax = sz_right, ymin = sz_bot, ymax = sz_top),
              col = "black", linetype = "dashed", alpha = 0) +
    geom_segment(aes(x = plate_left, xend = plate_right, y = 0, yend = 0)) +
    ggtitle(paste("Pitch Type:", next_type))
  plot(plt)
}
```


```{r save_res}
all_pitches <- all_pitches %>%
  mutate(two_strikes = (strikes == 2)) %>%
  left_join(looks, by = c("bin_ID", "two_strikes")) %>%
  left_join(swings, by = c("bin_ID", "pitch_type", "balls", "strikes")) %>%
  left_join(contact, by = c("bin_ID", "pitch_type")) %>%
  left_join(fouls, by = c("bin_ID", "pitch_type")) %>%
  left_join(barrels, by = c("bin_ID", "pitch_type")) %>%
  select(-two_strikes)

saveRDS(all_pitches, "output/mlb_wprobs.rds")
```