---
title: "Preprocessing"
author: "Jason Willwerscheid"
date: "2/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Scrape data

I used the `baseballr` package to scrape Statcast data from seasons 2017-2019. Since Statcast will only return a few days' worth of data at a time, a loop is required. For example, I scraped the 2017 data as follows:

```{r scrape, eval = FALSE}
# Not run:
season_begin <- as.Date("2017-04-02")
season_end <- as.Date("2017-11-01")

dat <- tibble()
start_date <- season_begin
while (start_date <= season_end) {
  cat("Start Date:", format(start_date, "%m-%d-%y"), "\n")
  dat <- dat %>%
    bind_rows(baseballr::scrape_statcast_savant_batter_all(start_date = start_date,
                                                           end_date = start_date + 6))
  start_date <- start_date + 7
}
saveRDS(dat, "data/mlb2017data.rds")
```

## Remove pitchers who hit and hitters who pitch

Since I want to capture average position players' behavior when facing average pitchers, I remove plate appearances where a pitcher is at the plate or where a position player is called in to pitch. I make specific exceptions for two-way players Shohei Ohtani and Michael Lorenzen.

```{r batvpit}
suppressMessages(library(tidyverse))

all_pitches <- readRDS("data/mlb2019data.rds") %>%
  bind_rows(readRDS("data/mlb2018data.rds")) %>%
  bind_rows(readRDS("data/mlb2017data.rds"))

pitchers <- all_pitches %>%
  group_by(pitcher) %>%
  summarize(n = n())
batters <- all_pitches %>%
  group_by(batter) %>%
  summarize(n = n())

# A player is defined as a pitcher if they throw more pitches than they face.
pitchers <- pitchers %>%
  left_join(batters, by = c("pitcher" = "batter"), suffix = c("_thrown", "_faced")) %>%
  filter(is.na(n_faced) | n_faced < n_thrown) %>% 
  pull(pitcher)

two_way <- unique(all_pitches$batter[all_pitches$player_name 
                                     %in% c("Shohei Ohtani", "Michael Lorenzen")])

prev_nrow <- nrow(all_pitches)

all_pitches <- all_pitches %>%
  filter(pitcher %in% pitchers, 
         batter %in% two_way | !(batter %in% pitchers))

# What percentage of pitches have I removed?
paste0(round(100 * (prev_nrow - nrow(all_pitches)) / prev_nrow, 1), 
       "% of pitches removed")
```

## Remove doubleheaders

To get count-specific wOBA values, I need to be able to determine the outcome of each plate appearance (PA). One problem with the Statcast data is that there's not a unique PA ID, and there's not a straightforward way to assign one. The best solution I came up with was to group by game date, home team, and at-bat number, but this method is unable to distinguish between games played in a doubleheader. I thought about additionally grouping by batter and pitcher, but PAs in which a pitching change occurred mid-at-bat would pose problems, and there is a small chance that the same batter will face the same pitcher in the same at-bat number in subsequent games of a doubleheader. Thus I elected to remove doubleheaders from the data.

```{r doubleheaders}
prev_nrow <- nrow(all_pitches)

PA_df <- all_pitches %>%
  group_by(game_date, home_team, at_bat_number) %>%
  summarize(is_doubleheader = (sum(balls == 0 & strikes == 0) == 2)) %>%
  ungroup() %>%
  mutate(PA_ID = row_number())

all_pitches <- all_pitches %>%
  left_join(PA_df, by = c("game_date", "home_team", "at_bat_number")) %>%
  filter(!is_doubleheader)

# Get the wOBA values for each PA outcome.
all_pitches <- all_pitches %>%
  mutate_at(vars(woba_value, woba_denom), ~ replace_na(., 0))
PA_woba <- all_pitches %>%
  group_by(PA_ID) %>%
  summarize(pa_woba_denom = max(woba_denom), pa_woba_value = max(woba_value)) %>%
  ungroup()
all_pitches <- all_pitches %>%
  left_join(PA_woba, by = "PA_ID")

paste0(round(100 * (prev_nrow - nrow(all_pitches)) / prev_nrow, 1), 
       "% of pitches removed")
```

## Remove rare events

I remove pitchouts, HBPs, and batter and catcher interference. There are also a small number of PAs where the count reaches four balls. I don't know whether these are errors in data entry or whether the umpires lost track of the count.

```{r outliers}
prev_nrow <- nrow(all_pitches)

all_pitches <- all_pitches %>%
  filter(balls < 4, !(description == "pitchout")) %>%
  filter(!(events %in% c("hit_by_pitch", "batter_interference", "catcher_interf")))

paste0(round(100 * (prev_nrow - nrow(all_pitches)) / prev_nrow, 1), 
       "% of pitches removed")
```

## Remove rare pitch types

I remove pitchouts (apparently, filtering by event isn't enough), eephus pitches, knuckleballs, and screwballs. I relabel knuckle curves as curveballs, since the pitch movement is so similar. For the same reason, sinkers can be grouped with two-seamers, and for better or worse, I group cutters with sliders and splitters with changeups. This leaves five loosely defined pitch types: four-seamers, two-seamers, sliders, changeups, and curveballs.

```{r pitch_type}
prev_nrow <- nrow(all_pitches)

all_pitches <- all_pitches %>%
  mutate(pitch_type = case_when(pitch_type == "KC" ~ "CU", 
                                pitch_type == "SI" ~ "FT",
                                pitch_type == "FC" ~ "SL",
                                pitch_type == "FS" ~ "CH",
                                TRUE ~ pitch_type)) %>%
  filter(pitch_type %in% c("CH", "CU", "FF", "FT", "SL"))
  
paste0(round(100 * (prev_nrow - nrow(all_pitches)) / prev_nrow, 1), 
       "% of pitches removed")  
```

## Remove missing data

I need to remove pitches where pitch location data is missing. The event frequencies (ball, blocked ball, called strike, etc.) are very similar between pitches that include this data and those that don't, so I think it's safe to assume that this data is missing at random.

```{r missing}
prev_nrow <- nrow(all_pitches)

all_pitches <- all_pitches %>%
  filter(!is.na(plate_x) & !is.na(plate_z))

paste0(round(100 * (prev_nrow - nrow(all_pitches)) / prev_nrow, 1), 
       "% of pitches removed")
```

## Flip location data for left-handed hitters

Since right-handed hitters' preferred pitch zones are mirror images of left-handed hitters', I flip the $x$-axis for pitches to lefties. This transformation eliminates the need to condition on handedness in subsequent computations.

```{r flip}
all_pitches <- all_pitches %>%
  mutate(plate_x = ifelse(stand == "L", -plate_x, plate_x))
```

## Bin location data

This final step is the most sensitive part of the preprocessing. I'd like to bin the pitch locations to get nonparametric estimates of swing rates and contact rates. I want to use as fine a grid as possible while retaining enough data points per bin to get reasonably accurate estimates. After some fiddling, I settled on bins with widths and heights of 0.2 feet (2.4 inches). I discarded pitches that landed in very sparsely populated bins: these are pitches that are located far off the plate and are easy to lay off of, so they don't tell us much about plate discipline.

In the plot below, the dashed lines indicate the [theoretical strike zone](https://www.baseballprospectus.com/news/article/37347/robo-strike-zone-not-simple-think/). The green bins are the ones that I've retained.

```{r loc}
prev_nrow <- nrow(all_pitches)

all_pitches <- all_pitches %>%
  mutate(plate_x = round(plate_x / 2, 1) * 2,
         plate_z = round(plate_z / 2, 1) * 2) %>%
  group_by(plate_x, plate_z)

plate_left <- -8.5 / 12
plate_right <- 8.5 / 12
sz_left <- plate_left - 2.94 / 2 / 12
sz_right <- plate_right + 2.94 / 2 / 12
sz_bot <- 15.53 / 12
sz_top <- 42.47 / 12

ggplot(all_pitches %>% summarize(n = n()), 
       aes(plate_x, plate_z, fill = n >= 1000, alpha = n)) +
  geom_tile() + 
  xlim(c(-5, 5)) +
  ylim(c(-3, 7)) +
  theme_minimal() +
  guides(fill = FALSE, alpha = FALSE) +
  scale_fill_manual(values = c("grey", "dark green")) +
  geom_rect(aes(xmin = sz_left, xmax = sz_right, ymin = sz_bot, ymax = sz_top),
                col = "black", linetype = "dashed", alpha = 0) +
  geom_segment(aes(x = plate_left, xend = plate_right, y = 0, yend = 0))
  
all_pitches <- all_pitches %>%
  filter(length(plate_x) >= 1000) %>%
  ungroup()

paste0(round(100 * (prev_nrow - nrow(all_pitches)) / prev_nrow, 1), 
       "% of pitches removed")
```

## Refactor variables

Finally, I refactor some variables, remove others, and save the results to file.

```{r save_data}
all_pitches <- all_pitches %>%
  mutate(swing = 1L - (description %in% c("ball", "blocked_ball", "called_strike")),
         contact = 1L * (description == "foul" | str_starts(description, "hit_into_play")),
         foul = 1L * (description %in% c("foul", "foul_pitchout")),
         called_strike = 1L * (description == "called_strike")) %>%
  select(PA_ID, pa_woba_value, pa_woba_denom, game_year, player_name, balls, strikes, 
         pitch_type, plate_x, plate_z, swing, contact, foul, called_strike, 
         woba_value, woba_denom)

saveRDS(all_pitches, "output/mlb_preproc.rds")
```